---
title: "BayesTME-LDA Analysis on Cohort Data"
author: "Joaquim Teixeira"
date: "2023-01-18"
output: html_document
---

```{r setup, include=FALSE}
library(readr)
library(stringr)
library(ncdf4)
library(data.table)
library(tidyr)
library(ggplot2)
```

```{r functions, include=FALSE}
jet.colors <-
  colorRampPalette(c("#00007F", "blue", "#007FFF", "cyan",
                     "#7FFF7F", "yellow", "#FF7F00", "red", "#7F0000"))

get_coords_cdata=function (x)
{
  v=str_split(x,'x',simplify = T)
  v=data.frame (v)
  
  v$X2=as.numeric(as.character(v$X2))
  v$X1=as.numeric(as.character(v$X1))
  v=data.frame (v)
  return (v)

}
get_nc_list=function (ncf)
{
  nc=nc_open (ncf)
  x=list()
  for (n in names (nc$var))
  {
    x[[n]]=ncvar_get(nc,n)
  }
  return (x)
  nc_close (nc)
}



analysis_slice=function (f_coords, f_btme, f_lda)
{
  dat=read.table(f_coords)
  cds=get_coords_cdata(dat[1,-1])
  lda=get_nc_list(f_lda)
  btme=get_nc_list(f_btme)
  lda$gene_probs=lda$components_/rowSums(lda$components_)
  #match=match_clusters(cds, apply  (t(btme$cell_num_trace)[,-1],1,which.max), apply  (t(lda$transform),1,which.max))
  probs_lda=data.frame (cds, t(lda$transform))
  names (probs_lda)=c('x','y',paste ('type_',1:6,sep=''))
  probs_btme=data.frame (cds, t(btme$cell_prob_trace[-1,]))
  names (probs_btme)=c('x','y',paste ('type_',1:6,sep=''))
  probs_lda=pivot_longer(probs_lda,cols=starts_with('cluster'))
  probs_btme=pivot_longer(probs_btme,cols=starts_with('cluster'))
  probs_lda$type='LDA'
  probs_btme$type='BayesTME'
  probs=rbind (probs_lda, probs_btme)
  g1= ggplot (probs, aes (x,y, fill=value))+geom_tile(color='black')+theme_bw()+facet_grid(name~type)+scale_fill_gradientn(colors=jet.colors (11),breaks =seq(0,1,length.out=11))+labs (x='',y='',fill='Spot Probability per Cell Type')  
  dfl=list()
  for (i in 1:6)
  {
    dfl[[i]]=data.frame (lda$gene_probs[btme$marker_genes[,i]+1,],cell_type=i,gene=lda$gene_names[btme$marker_genes[,i]+1])
  }
  df=rbindlist(dfl)
  names (df)=c(1:6, 'cell_type','gene')
  df=pivot_longer(df, cols=1:6)
  df$name=as.numeric(as.character(df$name))
  df$cell_type=paste('BayesTME_cell_type_',df$cell_type,sep='')
  g2=ggplot (df, aes (name,value, color=gene))+geom_line()+facet_wrap(~cell_type)+theme_bw()+theme(legend.position = 'bottom')+scale_x_continuous(breaks=seq(1,6,1))+labs (x='LDA Cell Type',y='LDA Gene Expression Probability\n for BayesTME Marker Genes',color='BayesTME Marker Genes')
  return (list(g1,g2,lda,btme))
}

analysis_lda_patient=function (f_coords, f_btme, f_lda_slice,f_lda_patient, f_clust,indvals)
{
  dat=read.table(f_coords)
  cds=get_coords_cdata(dat[1,-1])
  lda=get_nc_list(f_lda_slice)
  btme=get_nc_list(f_btme)
  lda$gene_probs=lda$components_/rowSums(lda$components_)
  probs_lda=data.frame (cds, t(lda$transform))
  names (probs_lda)=c('x','y',paste ('type_',1:6,sep=''))
  probs_btme=data.frame (cds, t(btme$cell_prob_trace[-1,]))
  names (probs_btme)=c('x','y',paste ('type_',1:6,sep=''))
  probs_lda=pivot_longer(probs_lda,cols=starts_with('type'))
  probs_btme=pivot_longer(probs_btme,cols=starts_with('type'))
  probs_lda$type='2-LDA_slice'
  probs_btme$type='1-BayesTME'
  hcldf=read.table (f_clust)
  probs_hclust=data.frame (x=hcldf$x,y=hcldf$y, name=paste ('type_',hcldf$label+1,sep=''))  
  probs_hclust$value=1
  probs_hclust$type='4-KNN'
  lda_pat=get_nc_list(f_lda_patient)
  probs_lda_patient=data.frame (cds, t(lda_pat$transform)[(indvals[1]:indvals[2]),])
  names (probs_lda_patient)=c('x','y',paste ('type_',1:6,sep=''))
  probs_lda_patient=pivot_longer(probs_lda_patient,cols=starts_with('type'))
  probs_lda_patient$type='3-LDA_patient'
  probs=rbind (probs_lda, probs_btme,probs_hclust,probs_lda_patient)
  
  
  g1= ggplot (probs, aes (x,y, fill=value))+geom_tile(color='black')+theme_bw()+facet_grid(name~type)+scale_fill_gradientn(colors=jet.colors (11),breaks =seq(0,1,length.out=11))+labs (x='',y='',fill='Spot Probability by\n Cell Type')  
  dfl=list()
  for (i in 1:6)
  {
    dfl[[i]]=data.frame (lda$gene_probs[btme$marker_genes[,i]+1,],cell_type=i,gene=lda$gene_names[btme$marker_genes[,i]+1])
  }
  df=rbindlist(dfl)
  names (df)=c(1:6, 'cell_type','gene')
  df=pivot_longer(df, cols=1:6)
  df$name=as.numeric(as.character(df$name))
  df$cell_type=paste('BayesTME_cell_type_',df$cell_type,sep='')
  g2=ggplot (df, aes (name,value, color=gene))+geom_line()+facet_wrap(~cell_type)+theme_bw()+theme(legend.position = 'bottom')+scale_x_continuous(breaks=seq(1,6,1))+labs (x='LDA Cell Type',y='LDA Gene Expression Probability\n for BayesTME Marker Genes',color='BayesTME Marker Genes')
  return (list(g1,g2,lda,btme))
}


```

```{r files, include=FALSE}
flc=list.files ('/Users/joaquimteixeira/GSR_22_Tansey_Banerjee/teixeira_code/flipped_cohort_data',full.names = T)
flb=list.files('/Users/joaquimteixeira/GSR_22_Tansey_Banerjee/teixeira_code/deconvolution_results_1000_sims',full.names = T)
fll=list.files('/Users/joaquimteixeira/GSR_22_Tansey_Banerjee/teixeira_code/ldas_by_slice',full.names = T)
fllp=list.files('/Users/joaquimteixeira/GSR_22_Tansey_Banerjee/teixeira_code/ldas_by_patient',full.names = T)
fllp=c(rep(fllp[1],6),rep(fllp[2],6),rep(fllp[3],6),rep(fllp[4],6),rep(fllp[5],3),rep(fllp[6],3),rep(fllp[7],3))

flh=list.files ('/Users/joaquimteixeira/GSR_22_Tansey_Banerjee/her2st-master/res/ST-cluster/lbl/',full.names = T)[seq(1,72,by=2)]
inds=read_csv ('/Users/joaquimteixeira/GSR_22_Tansey_Banerjee/teixeira_code/dim_per_slice.csv')
inds=as.matrix(inds[,-1])
patients=substr(flc,81,81)

```

We looked at a comparison between a few ST analyses on the HER2-positive breast tumors provided in Andersson et al. (2021). The data consists of 7 patients each with 3-6 slices. So far, this analysis is only on patient 'A'. The following analyses were performed:

1. BayesTME applied on each slice. We enforce 6 cell types and lambda of 10000. We sample 1000 times, with 2000 burn in steps and a thinning of 5. 

2. Latent Direchlet Allocation (LDA) applied  on each slice, based on sklearn.decomposition.LatentDirichletAllocation. We enforce 6 cell types.

3. LDA applied on each patient across all slices. Again, we enforce 6 cell types.

4. The baseline analysis performed by Andersson et al. (2021). K-Nearest neighbors clustering  is performed on gene expression at the patient+cohort level. This was used to set the number of cell types in each slice. 

This set of figures shows the spot probability by cell type for each of the 4 methods. 

```{r analysis, echo=FALSE}
vals=list()

for (i in 1:6)
{
  vals[[i]]=analysis_lda_patient(flc[i],flb[i],fll[i],fllp[i],flh[i],inds[i,])
}
```




```{r plots_1,echo=FALSE}
for (i in 1:length(vals))
{
  print(vals[[i]][[1]]+ggtitle (paste('Patient: ',patients[i],'; Slice: ',i,sep='')))
}

```

This set of figures considers only the BayesTME and slice-based LDA. For each BayesTME cell type, we compute the 5 highest marker genes. We then analyze  gene-topic probability from the slice-based LDA, to assess whether both methods have similar gene-latent space relationships.

```{r plots_2,echo=FALSE}
for (i in 1:length(vals))
{
  print(vals[[i]][[2]]+ggtitle (paste('Patient: ',patients[i],'; Slice: ',i,sep='')))
}

```

